---
title: "An introduction to loadr"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{An introduction to loadr}
output: knitr:::html_vignette
---

# An introduction to loadr

## Motivation

In a large, complex R analysis, I often find myself with dozens of resource data objects that I use as a reference for my computation. These are usually some kind of large dataset that I read into R and then use as reference data. As an example, I have a file with a big matrix with values from chromatin accessibility for 120 different cell-types across about a million different genomic regions (so, a 120-by-1 million matrix). I read this into my R sessions occasionally when I want to see, for example, how a new set of regions looks across that data resource. I'm not really changing the resource data, I'm just using it as a reference. I may load dozens of these kinds of resources into a single R session, and I want to re-use them across multiple scripts and multiple sessions, so I find myself reading them in a lot. I also start writing functions around these kinds of resources.

Given this complexity, sometimes it gets confusing for me to keep track of all the variables I keep in my primary R workspace, so I started to investigate how I could use R `environments` to declutter things. I discovered that I could load these kinds of "shared data reference" sets into a separate R environment, let's call it `SV` for "Shared Variables" -- and then they don't clutter my primary R workspace and I can still access them with `SV$variable`. This cleaned things up a bit for me and so I've developed a series of functions that make it really easy to interact with reference data like this kept in its own separate environment. That's what `loadr` is.

Before we start loading data, it's important to tell `loadr` the name of the environment where I want to store these reference objects. `loadr` uses a global variable (`LOADRENV`) to keep track of this environment name and provides a getter/setter function (`loadrEnv()`) to change or retrieve this. By default, we'll use an environment named `SV`. To make it explicit, just do this:

```{r Try it out}
library(loadr)
loadrEnv("SV")
```

(This is optional, `SV` is the default environment name. You can also use this to name your shared variable environment something else). Now, let's generate some random data and load it into that environment with `loadr` functions:

```{r}
eload(list(sampleData = rnorm(1e7, 0,1)))
```

Notice that the `eload` function takes a *named list*. The name of item in the list becomes the variable name of the object. If you now look at the R objects in your workspace, you won't see an object named `sampleData`, but you will see one named `SV`, which is a object of type `environment`:

```{r}
ls()
class(SV)
```

The `sampleData` object has been put in the `SV` environment:

```{r}
ls(envir=SV)
```

We can get that `sampleData` object back like this:

```{r}
head(SV$sampleData)
```

If you have multiple objects to load at the same time, no problem:

```{r}
eload(list(x=5, y=7))
```

The function has notified us that the `x` and `y` objects are newly loaded in to the shared variable environment. What happens if we try to load objects with the same name? They will get updated:

```{r}
eload(list(x=526, y=234))
SV$x
SV$y
```

In practice, I write functions that load up data and then I want to load up the results of these functions. Let's define a function that will read data from a (potentially large) file:

```{r}
loadMyData = function() {
	filePath = system.file("extdata", "mydata.txt", package="loadr")
	myData = read.table(filePath)
	return(myData)
}
```

```{r}
eload(list(myvector=loadMyData()))
SV$myvector
```

## Loading local variables or variably named variables with `vload`

What if I just want to load a local R variable, `myLocalVar`, under its current name?  `eload(myLocalVar)` doesn't work because there is no named list of variables, so that will fail. It's possibe but a bit cumbersome to use `eload(list(myLocalVar=myLocalVar))`.

Here's where an alias called `vload()` can help:

```{r}
myLocalVar = 15
vload(myLocalVar)

```

`vload` can accept any number of variables:

```{r}
myLocalVar2 = 22
myLocalVar3 = 31
vload(myLocalVar2, myLocalVar3)
```

You can also use `vload` with a special named argument `varNames` to assign the objects to variably named variables. For example:

```{r}
myLocalVar4 = 15
vload(myLocalVar2, myLocalVar3, varNames=c("var1", "var2"))
```

To see how this could be useful, let's imagine our data loading function depends on some variable which is script dependent, like the reference genome assembly:


```{r}
loadMyData = function(genome) {
	filePath = system.file("extdata", "mydata.txt", package="loadr")
	myData = read.table(filePath)
	return(myData)
}
```

Let's load up some data from this function for a few different genomes at the same time:

```{r}
genome="hg19"
vload(loadMyData(genome), varNames=paste0("myvector_", genome))

genome="mm10"
vload(loadMyData(genome), varNames=paste0("myvector_", genome))
```

An easier way to accomplish the same thing that will be to just `lapply` across our list of genomes:

```{r}
genomes=c("hg19", "mm10")
res = lapply(genomes, function(genome) { 
	vload(loadMyData(genome), varNames=paste0("myvector_", genome))
	})
```